Apache Flink\texttrademark: Unified Stream and Batch Processing \\ in a Single Engine

### 2015-11-25 batch.tex
Reflects the google doc. Further changes will be done in latex.

A bounded data set is a special case of an unbounded data stream. Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by Flink’s features that we presented above. However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and iii) programs that process bounded data sets are amenable to additional optimizations, more efficient book keeping for fault tolerance, and staged scheduling. .

\vspace{2mm}
\noindent Flink approaches batch processing as follows:
\begin{itemize}
	\item Batch computations are executed by the same runtime as streaming computations. The runtime executable may be parameterized with blocked data streams to break up large computations in isolated stages that are scheduled successively.
	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).
	\item Blocking operators (e.g., joins and sorts) are simply operator implementations that happen to block until they have consumed their entire input. The runtime is not aware of whether an operator is blocking or not. These operators use managed memory provided by Flink (either on- or off- the JVM heap), and can spill to disk if their inputs exceed their memory bounds.
	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc)
	\item A query optimization layer transforms a DataSet program to an efficient executable.
\end{itemize}

\para{Query Optimization.} Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation. However, the arbitrary UDF-heavy DAGs that make up Flink’s dataflow programs do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer. For the same reason, cardinality and cost estimation methods are equally difficult to employ. Flink’s runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations. Flink’s optimizer enumerates different physical plans based on the concept of interesting properties propagation \cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans. The cost includes network/disk I/O and CPU cost. To overcome the cardinality estimation issues that were mentioned earlier,  Flink’s optimizer uses hints that are provided by the programmer.

\para{Memory Management.} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into a  memory segments. Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed. To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.  By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and use cache-efficient and robust algorithms [Graefe join paper] that scale gracefully under memory pressure.

\para{Native Iterations on top of Dataflows.}  Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program. The approach, implemented in Naiad \cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations. A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration \cite{low2012distributed}.

@@

>>>>

A bounded data set is a special case of an unbounded data stream. Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by Flink’s features that we presented above. However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and iii) programs that process bounded data sets are amenable to additional optimizations, more efficient book keeping for fault tolerance, and staged scheduling.

\vspace{2mm}
\noindent Flink approaches batch processing as follows:\vspace{-3mm}
\begin{itemize}
	\item Batch computations are executed by the same runtime as streaming computations. The runtime executable may be parameterized with blocked data streams to break up large computations in isolated stages that are scheduled successively. \vspace{-3mm}
	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).\vspace{-3mm}
	\item Blocking operators (e.g., joins and sorts) are simply operator implementations that happen to block until they have consumed their entire input. The runtime is not aware of whether an operator is blocking or not. These operators use managed memory provided by Flink (either on- or off- the JVM heap), and can spill to disk if their inputs exceed their memory bounds.\vspace{-3mm}
	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc).\vspace{-3mm}
	\item A query optimization layer transforms a DataSet program to an efficient executable.\vspace{-3mm}
\end{itemize}

\subsection{Query Optimization}

Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation. However, the arbitrary UDF-heavy DAGs that make up Flink’s dataflow programs do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer. For the same reason, cardinality and cost estimation methods are equally difficult to employ. Flink’s runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations. Flink’s optimizer enumerates different physical plans based on the concept of interesting properties propagation \cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans. The cost includes network/disk I/O and CPU cost. To overcome the cardinality estimation issues that were mentioned earlier,  Flink’s optimizer uses hints that are provided by the programmer.

\subsection{Memory Management} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into a  memory segments. Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed. To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.  By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and use cache-efficient and robust algorithms [Graefe join paper] that scale gracefully under memory pressure.

\subsection{Native Iterations on top of Dataflows}
\label{sec:batch-iterations}
Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program. The approach, implemented in Naiad \cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations. A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration \cite{low2012distributed}.

@@

\subsection{Fault Tolerance}
\label{sec:batch-fault-tolerance}
\TODO{We need to simply say that we use the blocking data exchange to materialize intermediate results and redeploy in a spark fashion.}

### 2015-11-25 exec-model.tex
Reflects the google doc. Further changes will be done in latex.

Although users can write Flink programs using a multitude of APIs, all Flink programs eventually compile down to a common representation: the dataflow graph\footnote{Internally called JobGraph}. The dataflow graph is executed by Flink's runtime engine, the common layer underneath both the batch processing (DataSet) and stream processing (DataStream) APIs.

\subsection{Dataflow Graphs}
The dataflow graph, as depicted in \autoref{fig:dataflow}, is a directed acyclic graph (DAG) that consists of (i) stateful operators, and (ii) data streams that represent data produced by an operator and are available for consumption by operators. Since dataflow graphs are executed in a data-parallel fashion, operators are parallelized into one or more parallel instances called \emph{subtasks} and streams are split into one or more \emph{stream partitions} (one partition per producing subtask).

\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figs/dataflow}
\vspace{-3mm}
\caption{A simple dataflow graph comprising stateful operators (sources, sinks and operators), and intermediate streams (transient and materialized).}
\label{fig:dataflow}
\end{figure}

The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions, etc. Many of these operators are implementations of textbook versions of well known algorithms. \autoref{sec:streaming} provides details on the implementation of windowing operators. Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, or merge.

\subsection{Data Exchange through Intermediate Data Streams}
Flink's intermediate data streams are the core abstraction for data exchange between operators. The notion of intemediate data streams in Flink is broader than the traditional notion of a continuous transient stream. The particular behavior of a data stream is parameterized by the higher layers in Flink (e.g., the program optimizer used by the DataSet API).

\para{Pipelined Data Exchange.} Pipelined streams exchange data between concurrently running producers and consumers resulting in pipelined execution. Synchronous streams propagate back pressure from consumer to producer (modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations). Flink uses these streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible.

\para{Blocking Data Exchange.} Blocking streams are applicable to bounded data streams. The decoupled stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages. Blocking streams naturally require more memory, frequently spill to secondary storage, and do not propagate backpressure. They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as joins, can cause distributed deadlocks.

@@

@@

\para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events. These are special events injected in the data stream by producing operators and are delivered in-order along with all other data records and events within a stream partition. The receiving operators react to these events by performing certain actions upon their arrival. Flink uses control records in multiple mechanisms, such as: 
\begin{itemize}
\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (Section XCY)
\item To signal the progress of event time within a stream partition (Section XYZ). 
\item Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows, to signal that a stream partition has reached the end of a superstep.
\end{itemize}

\subsection{Iterative Dataflows}
{\color{red} Should we have a few sentences here? Would describe that cyclic flows enable iterative processing as BSP/SSP in batch [iterations paper, euranova SSP on Flink paper] and simply for streaming programs with feedback.}

@@

\subsection{Fault Tolerance}
Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution. The general assumption the system makes to effectively provide these guarantees is that the data sources are persistent and replayable. Examples of such sources are files and durable message queues (e.g. Apache Kafka). In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators. We provide more details about the approach we adopted to achieve fault tolerance for the DataSet and DataStream models respectively in sections 4.3 and 5.

>>>>

Although users can write Flink programs using a multitude of APIs, all Flink programs eventually compile down to a common representation: the dataflow graph. The dataflow graph is executed by Flink's runtime engine, the common layer underneath both the batch processing (DataSet) and stream processing (DataStream) APIs.

\subsection{Dataflow Graphs}
The dataflow graph, as depicted in \autoref{fig:dataflow}, is a directed acyclic graph (DAG) that consists of (i) stateful operators, and (ii) data streams that represent data produced by an operator and are available for consumption by operators. Since dataflow graphs are executed in a data-parallel fashion, operators are parallelized into one or more parallel instances called \emph{subtasks} and streams are split into one or more \emph{stream partitions} (one partition per producing subtask). 
The stateful operators (which may be stateless as a special case) implement all processing logic, e.g., filters, hash joins, stream window functions, etc. Many of these operators are implementations of textbook versions of well known algorithms. \autoref{sec:streaming} provides details on the implementation of windowing operators. Streams distribute data between producing and consuming operators in various patterns, such as point-to-point, broadcast, re-partition, fan-out, or merge.

\begin{figure}[t!]
\centering
\includegraphics[width=.7\textwidth]{figs/dataflow}
\vspace{-3mm}
\caption{A simple dataflow graph comprising stateful operators (sources, sinks and operators), and intermediate streams (transient and materialized).}
\label{fig:dataflow}
\end{figure}

\subsection{Data Exchange through Intermediate Data Streams}
\TODO{Asterios does not like the structure of this subsection. Minimal changes are needed but will be useful. for now it's a soup of ideas. Good news is, the content is already there.}

Flink's intermediate data streams are the core abstraction for data exchange between operators. The notion of intemediate data streams in Flink is broader than the traditional notion of a continuous transient stream. The particular behavior of a data stream is parameterized by the higher layers in Flink (e.g., the program optimizer used by the DataSet API).

\para{Pipelined and Blocking Data Exchange.} \emph{Pipelined streams} exchange data between concurrently running producers and consumers resulting in pipelined execution. Synchronous streams propagate back pressure from consumer to producer (modulo some elasticity via intermediate buffer pools, in order to compensate for short-term throughput fluctuations). Flink uses these streams for continuous streaming programs, as well as for many parts of batch dataflows, in order to avoid materialization when possible. \emph{Blocking streams}, on the other hand, are applicable to bounded data streams. The decoupled stream buffers all of the producing operator's data before making it available for consumption, thereby separating the producing and consuming operators into different execution stages. Blocking streams naturally require more memory, frequently spill to secondary storage, and do not propagate backpressure. They are used to isolate successive operators against each other (where desired) and in situations where plans with pipeline-breaking operators such as joins, can cause distributed deadlocks.

@@

\noindent \TODO{Do we need the one below? We will need more context if we decide to keep this here. Maybe move to streams?}

@@

\para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events. These are special events injected in the data stream by producing operators and are delivered in-order along with all other data records and events within a stream partition. The receiving operators react to these events by performing certain actions upon their arrival. Flink uses control records in multiple mechanisms, such as: \vspace{-2mm}
\begin{itemize}
\item Signaling the progress of event time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
\item Signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}). \vspace{-3mm}
\end{itemize}

\subsection{Iterative Dataflows}
\TODO{Should we have a few sentences here? Would describe that cyclic flows enable iterative processing as BSP/SSP in batch [iterations paper, euranova SSP on Flink paper] and simply for streaming programs with feedback. @Stephan, could you make a start here? Just to see how you would like to see it.}

@@

\subsection{Fault Tolerance}
Flink offers reliable execution with strict exactly-once-processing consistency guarantees and deals with failures via checkpointing and partial re-execution. The general assumption the system makes to effectively provide these guarantees is that the data sources are persistent and replayable. Examples of such sources are files and durable message queues (e.g. Apache Kafka). In practice, non-persistent sources can also be incorporated by keeping a look-ahead log within the state of the source operators. We provide more details about the approach we adopted to achieve fault tolerance for DataStreams and DataSets in sections \ref{sec:fault-tolerance} and \ref{sec:batch-fault-tolerance} respectively.

### 2015-11-25 intro.tex
Reflects the google doc. Further changes will be done in latex.

At the same time, Flink acknowledges that there is, and will be, a need for dedicated batch processing (dealing with static data sets). Complex queries over static data are still a good match for a batch processing abstraction. Furthermore, batch processing is still needed both for legacy implementations of streaming use cases, and for analysis applications where no efficient algorithms are yet known that perform this kind of processing on streaming data. Batch programs are special cases of streaming programs, where the stream is finite, and order/time of records does not matter (all records implicitly belong to one all-encompassing window). However, to support batch use cases with a competitive ease and performance, Flink has a specialized API for processing static data sets, uses specialized data structures and algorithms for the batch versions of operators like join or grouping, and uses dedicated scheduling strategies. The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning. 
Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing~\cite{CUSTOM:web/FlinkContrib}), and is used in production in several companies.

\noindent The contributions of this paper are the following:
\begin{itemize}
	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
	\vspace{-3mm}
	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault tolerant streaming dataflows (\autoref{sec:snapshots}).
	\vspace{-3mm}
	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:windows}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
\end{itemize}

>>>>

At the same time, Flink acknowledges that there is, and will be, a need for dedicated batch processing (dealing with static data sets). Complex queries over static data are still a good match for a batch processing abstraction. Furthermore, batch processing is still needed both for legacy implementations of streaming use cases, and for analysis applications where no efficient algorithms are yet known that perform this kind of processing on streaming data. Batch programs are special cases of streaming programs, where the stream is finite, and order/time of records does not matter (all records implicitly belong to one all-encompassing window). However, to support batch use cases with a competitive ease and performance, Flink has a specialized API for processing static data sets, uses specialized data structures and algorithms for the batch versions of operators like join or grouping, and uses dedicated scheduling strategies. The result is that Flink presents itself as a full-fledged and efficient batch processor on top of a streaming runtime, including libraries for graph analysis and machine learning. 
Originating from the Stratosphere project \cite{stratosphere}, Flink is a top-level project of the Apache Software Foundation that is developed and supported by a large and lively community (of more than 140 open source contributors at the time of this writing\footnote{For an update list, refer to \url{https://github.com/apache/flink/graphs/contributors}}), and is used in production in several companies.

\vspace{2mm}
\noindent The contributions of this paper are the following:\vspace{-2mm}
\begin{itemize}
	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
	\vspace{-3mm}
	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault tolerant streaming dataflows (\autoref{sec:execution}).
	\vspace{-3mm}
	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:streaming}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
\end{itemize}

### 2015-11-25 related.tex
Reflects the google doc. Further changes will be done in latex.

%!TEX root = paper.tex

\section{Related work}
\label{sec:related}
There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.

\para{Batch Processing.} Apache Hadoop\footnote{https://hadoop.apache.org/} is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}. Dryad~\cite{isard2007dryad}, introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer} which added a language and an SQL optimizer on top of it. Apache Tez\footnote{https://tez.apache.org/} can be seen as an open source implementation of the ideas proposed in Dryad. MPP databases \cite{dewitt1990gamma}, and recent open-source implementations~\cite{CUSTOM:web/Drill,kornacker2015impala} restrict their API to SQL variants. Similar to Flink, Apache Spark\footnote{spark.apache.org} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches. In contrast, Flink is the only system that 
\begin{inparaenum}[i)]
  \item supports an optimizer that can optimize DAG programs beyond SQL queries,
  \item is able to perform very efficient iterative processing natively,
  \item includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
\end{inparaenum}

\para{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams. Many of these systems are based on research in the database community in projects such as PSoup/Telegraph \cite{chandrasekaran2003psoup}, Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{gedik2008spade}. Most of the above systems are either
\begin{inparaenum}[i)]
  \item academic prototypes,
  \item closed-source commercial products, or 
  \item do not scale the computation horizontally on clusters of commodity servers.
\end{inparaenum}
Newer open source streaming systems that scale horizontally, such as Apache Storm\footnote{http://storm.apache.org/} and Apache Samza\footnote{http://samza.apache.org/}, provide low level APIs and offer only at-least-once and at-most-once guarantees. MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics. To the best of our knowledge, Flink is the only open-source project that
\begin{inparaenum}[i)]
  \item offers high level APIs and processing libraries,
  \item provides state management with exactly-once guarantees and
  \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.
\end{inparaenum}

>>>>

%!TEX root = paper.tex
\vspace{-4mm}
\section{Related work}
\vspace{-2mm}
\label{sec:related}
There is, by now, a wealth of engines for distributed batch and stream analytical processing. We categorise these systems below.

\para{Batch Processing.} Apache Hadoop is the most popular open source system for large-scale data analysis that is based on the MapReduce paradigm~\cite{DBLP:journals/cacm/DeanG08}. Dryad~\cite{isard2007dryad}, introduced embedded user-defined functions in general DAG-based dataflows and was used by SCOPE~\cite{scopeOptimizer} which added a language and an SQL optimizer on top of it. Apache Tez~\cite{saha2015apache} can be seen as an open source implementation of the ideas proposed in Dryad. MPP databases \cite{dewitt1990gamma}, and recent open-source implementations~\cite{CUSTOM:web/Drill,kornacker2015impala} restrict their API to SQL variants. Similar to Flink, Apache Spark \cite{DBLP:conf/hotcloud/ZahariaCFSS10} is a data processing framework that implements a DAG-based processing framework, provides an SQL optimizer, performs driver-based iterations, and treats stream computations as micro-batches. In contrast, Flink is the only system that 
\begin{inparaenum}[i)]
  \item supports an optimizer that can optimize DAG programs beyond SQL queries,
  \item is able to perform very efficient iterative processing natively,
  \item includes stream processing as a first-class citizen, enabling more complex use cases than micro-batches.
\end{inparaenum}

\para{Stream Processing.} There is a wealth of work on streaming systems, that includes commercial systems like StreamBase, Microsoft StreamInsight, and IBM InfoSphere Streams. Many of these systems are based on research in the database community in projects such as PSoup/Telegraph \cite{chandrasekaran2003psoup}, Aurora/Borealis~\cite{abadi2005design}, Stanford STREAM~\cite{arasu2004stream}, Trill~\cite{chandramouli2014trill} and IBM System S~\cite{gedik2008spade}. Most of the above systems are either
\begin{inparaenum}[i)]
  \item academic prototypes,
  \item closed-source commercial products, or 
  \item do not scale the computation horizontally on clusters of commodity servers.
\end{inparaenum}
Newer open source streaming systems that scale horizontally, such as Apache Storm and Apache Samza\footnote{Those can be found under \url{http://storm.apache.org/} and \url{http://samza.apache.org/} respectively.}, provide low level APIs and offer only at-least-once and at-most-once guarantees. MillWheel~\cite{akidau2013millwheel}, a closed-source system at Google provides exactly-once guarantees with low latency and is used by Google Dataflow~\cite{akidau2015dataflow} to implement  out-of-order stream analytics. To the best of our knowledge, Flink is the only open-source project that
\begin{inparaenum}[i)]
  \item offers high level APIs and processing libraries,
  \item provides state management with exactly-once guarantees and
  \item achieves high throughput and low latency, serving both batch and true streaming computations equally well.
\end{inparaenum}

### 2015-11-25 streaming.tex
Reflects the google doc. Further changes will be done in latex.

\section{Stream Analytics on Top of Dataflows}
\label{sec:windows}

@@

\subsection{The Notion of Time}
Flink distinguishes between two notions of time: i) event time, that  stands for the time that an event was generated at its origin (e.g., the time that a sensor or a mobile device generated a signal) and ii) processing time, that is the wall-clock time of the machine that is processing the data.

@@

Watermarks originate at the sources of the topology, where we can determine the time inherent in future elements accurately. The watermarks propagate from the sources throughout the other operators of the data flow. Operators decide how they react to watermarks. Simple operations such as map/filter just forward the watermarks they receive, while more complex operators that do calculation based on the watermarks (e.g., event time windows) first compute results triggered by a watermark and then forward it. If an operation has more than one input, the system only forwards the minimum of the incoming watermarks to the operator, ensuring correct results.

@@

\begin{lstlisting}[language=Scala]
stream
  .window(SlidingTimeWindows.of(Time.of(6, SECONDS), Time.of(2, SECONDS))
  .trigger(EventTimeTrigger.create())
\end{lstlisting}

@@

\begin{lstlisting}[language=Scala]
stream
  .window(GlobalWindow.create())
  .trigger(Count.of(1000))
  .evict(Count.of(100))
\end{lstlisting}

Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality [Dataflow].

\subsection{Stateful Stream Processing}
Simple programs in Flink’s DataStream API look like functional, side effect-free programs consisting of transformations on unbounded immutable collections. However, to cover the need for stateful operations, we incorporate explicit mutable state in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partionable dynamic state and its operations.  Flink’s checkpointing mechanism guarantees that any registered state is durably accessed exactly-once per operation via checkpointing.

@@

\begin{lstlisting}[language=Scala]
wordNums.keyBy(0).map(new StatefulAverage())

@@

\subsection{Fault Tolerance via Asynchronous Snapshots}
\label{sec:fault-tolerance}
The basic fault tolerance mechanism of Apache Flink for unbounded streams builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees. The possibly unbounded nature of a data stream in the model of the DataStream API makes recomputation upon recovery impractical, as possibly months of computation will need to be replayed considering a long running job. To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.

@@

\begin{figure}[ht]
	\centering  	
  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
  	\vspace{-4mm}
	\caption{Asynchronous Barrier Snapshotting.}
	\vspace{-2mm}
	\label{fig:FlinkStack}
\end{figure}

>>>>

\section{Stream Analytics on Top of Dataflows}
\label{sec:streaming}

@@

\subsection{The Notion of Time}
\label{sec:streaming-time}
Flink distinguishes between two notions of time: i) event time, that  stands for the time that an event was generated at its origin (e.g., the time that a sensor or a mobile device generated a signal) and ii) processing time, that is the wall-clock time of the machine that is processing the data.

@@

Watermarks originate at the sources of a topology, where we can determine the time inherent in future elements. The watermarks propagate from the sources throughout the other operators of the data flow. Operators decide how they react to watermarks. Simple operations such as map/filter just forward the watermarks they receive, while more complex operators that do calculation based on the watermarks (e.g., event time windows) first compute results triggered by a watermark and then forward it. If an operation has more than one input, the system only forwards the minimum of the incoming watermarks to the operator, ensuring correct results.

@@

\begin{lstlisting}[language=Java]
stream
  .window(SlidingTimeWindows.of(Time.of(6, SECONDS), Time.of(2, SECONDS))
  .trigger(EventTimeTrigger.create())
\end{lstlisting}

@@

\begin{lstlisting}[language=Java]
stream
  .window(GlobalWindow.create())
  .trigger(Count.of(1000))
  .evict(Count.of(100))
\end{lstlisting}

Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.

\subsection{Stateful Stream Processing}
\TODO{Asterios: This is very developer-heavy. We need to show the big picture. Long story short, the side-effect free functional view of the system changes with mutable state that is managed by the system. Why do we need state? What can be state? How is it implemented in Flink. @Paris, please answer these 3 questions in 2-3 paragraphs. The example below could also be missing.}

Simple programs in Flink’s DataStream API look like functional, side effect-free programs consisting of transformations on unbounded immutable collections. However, to cover the need for stateful operations, we incorporate explicit mutable state in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partionable dynamic state and its operations.  Flink’s checkpointing mechanism guarantees that any registered state is durably accessed exactly-once per operation via checkpointing.

@@

\begin{lstlisting}[language=Java]
wordNums.keyBy(0).map(new StatefulAverage())

@@

\subsection{Fault Tolerance via Asynchronous Snapshots}
\label{sec:fault-tolerance}
The fault tolerance mechanism of Apache Flink for unbounded streams builds on the notion of distributed consistent snapshots/checkpoints to achieve exactly-once-processing guarantees. The possibly unbounded nature of a data stream in the model of the DataStream API makes recomputation upon recovery impractical, as possibly months of computation will need to be replayed considering a long running job. To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.

@@

\begin{figure}[ht]
	\centering  	
  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
  	\vspace{-6mm}
	\caption{Asynchronous Barrier Snapshotting.}
	\vspace{-2mm}
	\label{fig:snapshots}
\end{figure}
